---
title: "Plotting mutational signatures with mSigPlot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting mutational signatures with mSigPlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 4,
  out.width = "100%"
)
```

## Introduction

mSigPlot creates publication-quality plots for mutational signatures and
mutational spectra. It supports SBS (single base substitution), DBS (doublet
base substitution), and indel catalogs across 10 classification systems.

```{r setup}
library(mSigPlot)
```

## Loading catalog data

mSigPlot works with numeric vectors, data frames, matrices, tibbles, and
data.tables. Here we load three example catalogs bundled with the package.

### SBS96 catalog

The 96-channel SBS catalog has one row per trinucleotide mutation context. The
bundled example contains mutation counts from four HepG2 samples.

```{r load-sbs96}
sbs96_file <- system.file("extdata", "sbs96_example.csv", package = "mSigPlot")
sbs96_df <- read.csv(sbs96_file)

# Build a named catalog: row names are the 96 canonical trinucleotide types
orders <- catalog_row_order()
catalog_sbs96 <- data.frame(
  sample1 = sbs96_df[, 3],
  row.names = orders$SBS96
)

head(catalog_sbs96)
```

### DBS78 catalog

The 78-channel DBS catalog covers all doublet base substitution classes.

```{r load-dbs78}
dbs78_file <- system.file("extdata", "dbs78_example.csv", package = "mSigPlot")
dbs78_df <- read.csv(dbs78_file)

catalog_dbs78 <- data.frame(
  sample1 = dbs78_df[, 3],
  row.names = paste0(dbs78_df$Ref, dbs78_df$Var)
)

head(catalog_dbs78)
```

### ID83 catalog (COSMIC signatures)

The 83-channel indel catalog uses the COSMIC classification. This file
contains COSMIC v3.5 reference signatures; each column is a signature.

```{r load-id83}
id83_file <- system.file("extdata", "id83_cosmic_v3.5.tsv", package = "mSigPlot")
id83_sigs <- read.table(id83_file, header = TRUE, sep = "\t",
                         row.names = 1, check.names = FALSE)

# Pick one signature to plot
catalog_id83 <- id83_sigs[, "ID1", drop = FALSE]
head(catalog_id83)
```

## Plotting individual catalogs

### SBS96

```{r plot-sbs96}
plot_SBS96(catalog_sbs96, plot_title = "HepG2 sample — SBS96")
```

### DBS78

```{r plot-dbs78}
plot_DBS78(catalog_dbs78, plot_title = "HepG2 sample — DBS78")
```

### ID83

```{r plot-id83}
plot_83(catalog_id83, plot_title = "COSMIC ID1 signature")
```

## Customizing plots

All plot functions accept parameters to control appearance:

```{r customize}
plot_SBS96(
  catalog_sbs96,
  plot_title = "Custom styling example",
  base_size   = 14,
  show_counts = TRUE,
  grid        = FALSE
)
```

## Auto-dispatch with plot_guess()

If you don't know (or don't want to specify) the catalog type,
`plot_guess()` detects it from the number of rows:

```{r plot-guess}
plot_guess(catalog_sbs96, plot_title = "Auto-detected SBS96")
```

## Multi-sample PDF export

For batch plotting, every plot function has a `_pdf()` variant that writes
a multi-page PDF with 5 plots per page. The auto-dispatch version is
`plot_guess_pdf()`:

```{r pdf-export, eval = FALSE}
# Build a multi-sample matrix (4 samples)
sbs96_mat <- as.matrix(sbs96_df[, 3:6])
rownames(sbs96_mat) <- orders$SBS96
colnames(sbs96_mat) <- paste0("Sample_", 1:4)

# Write to PDF — 5 plots per page
plot_guess_pdf(sbs96_mat, file.path(tempdir(), "sbs96_samples.pdf"))
```

## Combining plots

Since plot functions return ggplot objects, you can combine them using
packages like `patchwork`:

```{r patchwork, fig.height = 8}
if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p1 <- plot_SBS96(catalog_sbs96, plot_title = "SBS96")
  p2 <- plot_DBS78(catalog_dbs78, plot_title = "DBS78")
  p1 / p2
}
```
